---
title: "Practice: Vectors (part 1)"
subtitle: "Stat 133"
author: "Gaston Sanchez"
format: 
  html:
    toc: true
    number-sections: true
    theme: simplex
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


::: {.callout-note}
## Learning Objectives

- Work with vectors of different data types
- Understand the concept of _coercion_
- Understand the concept of _vectorization_
- Understand _recycling_ rules in R
:::



# Terrestrial Planets

In this module, we are going to use data from so-called __Terrestrial__
planets. These planets include Mercury, Venus, Earth, and Mars. They are called
like this because they are "Earth-like" planets in contrast to the __Jovian__
planets that involve planets similar to Jupiter (i.e. Jupiter, Saturn, Uranus 
and Neptune). The main characteristics of terrestrial planets is that they are 
relatively small in size and in mass, with a solid rocky surface, and metals 
deep in its interior.

| planet  | gravity | daylength | temp  | moons | haswater |
|:-------:|:-------:|:---------:|:-----:|:-----:|:--------:|
| Mercury |  3.7    | 4222.6    | 167   |   0   |  FALSE   |
| Venus   |  8.9    | 2802      | 464   |   0   |  FALSE   |
| Earth   |  9.8    |   24      |  15   |   1   |   TRUE   |
| Mars    |  3.7    |   24.7    | -65   |   2   |  FALSE   |



# Creating Vectors

The first step is to create vectors for each of the columns in the data table 
displayed above. For illustration purposes, we are going to assume the following
data types:

- `planet`: character vector

- `gravity`: real (i.e. double) vector ($m/s^2$)

- `daylength`: real (i.e. double) vector (hours)

- `temp`: integer vector (mean temperature in Celsius)

- `moons`: integer vector (number of moons)

- `haswater`: logical vector (whether has known bodies of liquid water on its surface)



## Combine function `c()`

The most common way to create an R vector is with the __combine__ function `c()`.
Here's an example:

```{r echo = TRUE}
jovians = c("Jupiter", "Saturn", "Uranus", "Neptune")
jovians
```


### Your Turn 

a) Create a character vector `planets` with the names of the Terrestrial planets.

```{r}
#| code-fold: true
#| code-summary: "Show answer"
planets = c("Mercury", "Venus", "Earth", "Mars")
```


b) Use the combine function `c()` to make vectors `gravity` and `daylength` for
the Terrestrial planets.

```{r}
#| code-fold: true
#| code-summary: "Show answer"
gravity = c(3.7, 8.9, 9.8, 3.7)

daylength = c(4222.6, 2802, 24, 24.7)
```


### Integer vectors

The creation of the temperature vector seems to be straightforward: 

```{r echo = TRUE}
temp <- c(167, 464, 15, -65)
temp
```

But there is a catch. The issue is that the way `temp` was created is as a 
vector of type `"double"` instead of type `"integer"` as required:

```{r echo = TRUE}
typeof(temp)
```

So how do you create integer vectors in R? You have to use a special notation
for integer numbers. Here's an example:

```{r echo = TRUE}
ints = c(2L, 4L, 6L)
ints
```

Notice how we append an upper case `L` at the end of every numeric value.
This is how you tell R to store such numbers as integers.


### Your Turn

a) Use the combine function to create integer vectors `temp` and `moons` for the 
Terrestrial planets. Inspect their data types, with `typeof()`, to confirm 
that they are integer vectors.

```{r}
#| code-fold: true
#| code-summary: "Show answer"
temp = c(167L, 464L, 15L, -65L)

moons = c(0L, 0L, 1L, 2L)
```


b) Use the combine function to create a logical vector (logical values are `TRUE`
and `FALSE`) for the variable `haswater`.

```{r}
#| code-fold: true
#| code-summary: "Show answer"
haswater = c(FALSE, FALSE, TRUE, FALSE)
```


-----


# Coercion

Now that we have various vectors---of different data types---to play with,
let's talk about __Coercion__ which is what R does when you try to combine 
distinct data types into a single vector.

We have an integer vector `ints`

```{r echo = TRUE}
ints
```

What happens if we combine `ints` with one or more double values?
What is the data type of the new vector `values`?

```{r echo = TRUE}
values = c(ints, 8, 10)
values

typeof(values)
```

Can you guess why `values` is not of type `"integer"` anymore?


## Your Turn

Inspect the data type of the following combination of vectors:

a) combine `planets` with `gravity`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(planets, gravity))
```


b) combine `planets` with `temp`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(planets, temp))
```


c) combine `planets` with `haswater`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(planets, haswater))
```


d) `gravity` with `daylength`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(gravity, daylength))
```


e) combine `gravity` with `temp`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(gravity, temp))
```


f) combine `temp` with `moons`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(temp, moons))
```


g) combine `temp` with `haswater`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
typeof(c(temp, haswater))
```

__Can you see a pattern?__


-----


# Vectorization

In addition to __coercion__, another fundamental concept to learn about R
vectors is that of __vectorization__. This is basically what R does when you
apply a function or an operation to all the elements of a vector in a 
simultaneous way.


## Vectorization Example

Here's an example. Let's bring back the integer vector `ints`, and suppose 
that we want to obtain the square root of all its elements. One option to do 
this is by taking the square root of each element in `ints`, one by
one---separately---using the `sqrt()` function:

```{r echo = TRUE}
sqrt(ints[1])
sqrt(ints[2])
sqrt(ints[3])
```

We haven't talked about this yet, but notice how you refer to the elements
in a vector by indicating their position: using square brackets `[ ]` with a 
numeric index for the position of the element you want to operate on.

Now, instead of having to repeat the same command three times, we can use the 
function `sqrt()` in a single call because it is a vectorized function. This 
means that `sqrt()` can compute the square root of all the elements in a vector 
simultaneously:

```{r echo = TRUE}
sqrt(ints)
```

Likewise, pretty much all arithmetic operators (addition, subtraction, 
multiplication, division, power, etc) are vectorized. For instance, say we 
want to add `c(1,2,3)` to `ints`, here's how to do it with vectorized code:

```{r echo = TRUE}
ints + c(1, 2, 3)
```



## Your Turn

a) Refer to the vector `gravity` and create a new vector `gravity_log` by taking
the logarithm, with `log()`, of the values in `gravity`

```{r}
#| code-fold: true
#| code-summary: "Show answer"
gravity_log = log(gravity)
```


b) Refer to the vectors `moons` and `haswater`. Try adding them and see what
happens. What is R doing in this case?

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
moons + haswater
```


c) Refer to the vectors `moons` and `haswater`. Try subtracting them, e.g. 
`moons - haswater`, and see what happens. What is R doing in this case?

```{r}
#| code-fold: true
#| code-summary: "Show answer"
#| eval: FALSE
moons - haswater
```


-----


# Recycling

Related to vectorization, there is another important concept called 
__recycling__ which has to do with what R does when you operate with two 
(or more) vectors of different length.


## Recycling Example

Consider the vectors `ints` and `values` which, by the way, are of different
length:

```{r echo = TRUE}
ints
values
```

What if we try to add `ints` and `values`? Is this possible?

```{r echo = TRUE}
ints + values
```

Yes, you can add two numeric vectors of different lengths such as `ints` plus
`values`. Notice though that R gives a __warning message__ along the lines of

> longer object length is not a multiple of shorter object length

This message tells you that the length of the longer vector, `values`, is not a 
multiple of the length of the shorter vector `ints`.

When computing `ints + values`, R is basically recycling or repeating some of 
the numbers in `ints` to match the length of `values`. To be more precise, 
here is what R is adding:

```
2L + 2 = 4
4L + 4 = 6
6L + 6 = 12
2L + 8 = 10
4L + 10 = 14
```

All the integer numbers come from `ints` whereas all the double numbers come
from `values`.


## Your Turn

a) Refer to the vector `daylength` (measured in hours) and create a new vector 
`dayminutes` in which the units are expressed in minutes instead of hours.

```{r}
#| code-fold: true
#| code-summary: "Show answer"
dayminutes = daylength * 60
```


b) Refer to the vector `temp` (measured in Celsius degrees) and create a new 
vector `temp2` in which the units are expressed in Fahrenheit degrees. 
The conversion factor from Celsius to Fahrenheit is:

$$
(1^{\circ}C × 9/5) + 32 = 33.8^{\circ}F
$$

```{r}
#| code-fold: true
#| code-summary: "Show answer"
temp2 = (temp * 9/5) + 32
```


c) Use the power operator `^` to raise `ints` to the values of `moons`:
e.g. `ints^moons`. If you get a warning, why is so? What is R doing behind 
this operation?

```{r}
#| code-fold: true
#| code-summary: "Show answer"
# R gives a warning because the longer vector "moons" is not a
# multiple of the shorter vector "ints"
```

