---
title: "North Atlantic Storms"
subtitle: "Mapping Hurricanes Data II"
author: "Gaston Sanchez"
format: 
  revealjs:
    slide-number: true
    show-slide-number: print
    footer: "[Return to Home Page](../..)"
    theme: ../custom.scss
---


```{r include=FALSE}
library(tidyverse)
library(leaflet)
library(sf)
library(rnaturalearth)
```


# Maps


## Leaflet: Web Interactive Maps

```{r}
# all tropycal cyclones in 1975
storms |>
  filter(year == 1975) |>
  leaflet() |>
  setView(lng = -50, lat = 30, zoom = 2) |>
  addProviderTiles(provider = 'CartoDB') |>
  addCircleMarkers(
    lng = ~long, 
    lat = ~lat,
    radius = 1)
```


## Leaflet

[Leaflet](https://leafletjs.com/) is a JavaScript library used to build 
interactive maps for websites and web mapping applications.

. . .

The R package `"leaflet"` lets you create and customize _Leaflet_ maps in R. 

. . .

These maps can be used directly from the R console, from RStudio, in Markdown 
documents (e.g. `qmd`, `Rmd`), and in Shiny applications.


## Basic Usage

You create a Leaflet map with these basic steps:

1) Create a map widget by calling `leaflet()`.

2) Add layers (i.e. features) to the map by using layer functions (e.g. 
`addTiles`, `addMarkers`, `addPolygons`) to modify the map widget.

3) Repeat step 2 as desired.

4) Print the map widget to display it.


## Leaflet: Basic Example

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
# view with marker at UC Berkeley
leaflet() |>
  addTiles() |> # default OpenStreetMap map tiles
  addMarkers(lng = -122.2579, lat = 37.8718, popup = "Berkeley")
```


## Basic Example: Map and Marker

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
# view with marker at Berkeley, CA
leaflet() |>
  addTiles() |>
  setView(lng = -122, lat = 38, zoom = 8) |>
  addMarkers(lng = -122.27, lat = 37.87, popup = "Berkeley")
```


## Basic Example: Map and Marker (cont'd)

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
# view with marker at Berkeley, CA
leaflet() |>
  addTiles() |>
  setView(lng = -100, lat = 37, zoom = 4) |>
  addMarkers(lng = -122.27, lat = 37.87, popup = "Berkeley")
```


## Map Tiles

Leaflet uses map tiles from:

- OpenStreetMap (OSM): <https://www.openstreetmap.org/>

    + OSM is simply the database with all the data in xml-vector format.

- CartoDB: <https://carto.com/basemaps/>


## Adding Data Objects

Both `leaflet()` and the map layer functions have an optional `data` parameter 
that is designed to receive spatial data in one of several forms:

- From base R:
    + lng/lat matrix
    + data frame with lng/lat columns
- From the `"sf"` package:
    + the data frame of class `"sf"`
- From the `"maps"` package
    + the data frame from returned from `map()`


## Example: adding `"sf"` data polygons

```{r}
usa_states = ne_states(
  country = "United States of America", 
  returnclass = "sf")

leaflet(data = usa_states) |> 
  addTiles() |>
  setView(lng = -90, lat = 40, zoom = 3) |>
  addPolygons(fillColor = rainbow(10), stroke = FALSE)
```



## Leaflet Maps

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
storms |>
  filter(year == 1975) |>
leaflet() |>
  setView(lng = -50, lat = 30, zoom = 3) |>
  addTiles() |>
  addProviderTiles(provider = "NASAGIBS.ViirsEarthAtNight2012") |>
  addCircleMarkers(
    lng = ~long, 
    lat = ~lat,
    radius = 1, 
    color = "#DDFF03")
```


## Leaflet Maps

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
storms |>
  filter(year == 1975) |>
leaflet() |>
  setView(lng = -50, lat = 30, zoom = 3) |>
  addTiles() |>
  addProviderTiles(provider = "NASAGIBS.ViirsEarthAtNight2012") |>
  addCircleMarkers(
    lng = ~long, 
    lat = ~lat,
    radius = 1, 
    color = "#DDFF03", 
    weight = ~wind/5)
```


