---
title: "North Atlantic Storms"
subtitle: "Mapping Hurricanes Data I"
author: "Gaston Sanchez"
format: 
  revealjs:
    slide-number: true
    show-slide-number: print
    footer: "[Return to Home Page](../..)"
    theme: ../custom.scss
---


```{r include=FALSE}
library(tidyverse)
library(sf)
library(rnaturalearth)
# library(rnaturalearthhires)
```


## Data `storms`

```{r}
storms
```


# Plotting Storms Paths


## Plot storms' paths

```{r}
#| code-fold: true
#| code-summary: View Plot Code
#| echo: true
storms |>
  filter(year == 1975) |>
  ggplot(aes(x = long, y = lat, group = name)) +
  geom_point() +
  labs(title = "North Atlantic Storms (1975)")
```


## Plot storms' paths

```{r}
#| code-fold: true
#| code-summary: View Plot Code
#| echo: true
storms |>
  filter(year == 1975) |>
  ggplot(aes(x = long, y = lat, group = name)) +
  geom_point() +
  geom_path() + 
  labs(title = "North Atlantic Storms (1975)")
```


## Plot storms' paths

```{r}
#| code-fold: true
#| code-summary: View Plot Code
#| echo: true
storms |>
  filter(year == 1975) |>
  ggplot(aes(x = long, y = lat, group = name, color = name)) +
  geom_point(size = 1) +
  geom_path() + 
  labs(title = "North Atlantic Storms (1975)")
```


## Plot storms' paths

```{r}
#| code-fold: true
#| code-summary: View Plot Code
#| echo: true
storms |>
  filter(year == 1975) |>
  ggplot(aes(x = long, y = lat, group = name, color = name)) +
  geom_point(size = 1) +
  geom_path(arrow = arrow(length = unit(0.1, "inches"))) + 
  labs(title = "North Atlantic Storms (1975)")
```


## Plot storms' paths

```{r}
#| code-fold: true
#| code-summary: View Plot Code
#| echo: true
storms |>
  filter(year == 1975) |>
  ggplot(aes(x = long, y = lat, group = name, color = name)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_path(arrow = arrow(length = unit(0.1, "inches"))) + 
  facet_wrap(~ month) +
  labs(title = "North Atlantic Storms (1975)") +
  theme_bw()
```


# Maps


## Wouldn't this be nice?

```{r}
#| echo: false
world_countries = ne_countries(returnclass = "sf")

storms75 = storms |> filter(year == 1975)

ggplot() +
  geom_sf(data = world_countries) +
  coord_sf(xlim = c(-110, 0), ylim = c(5, 65)) +
  geom_point(data = storms75, 
          aes(x = long, y = lat, group = name, color = name),
          size = 0.5, alpha = 0.5) +
  geom_path(data = storms75, 
          aes(x = long, y = lat, group = name, color = name),
          arrow = arrow(length = unit(0.1, "inches"))) + 
  theme(panel.background = element_blank(),
        axis.ticks = element_blank(), # hide tick marks
        axis.text = element_blank()) + # hide degree values of lat & lon
  labs(title = "North Atlantic Storms (1975)",
       x = "", 
       y = "")
```


## What about something like this?

```{r}
#| echo: false
world_countries = ne_countries(returnclass = "sf")

some_storms = storms |> filter(year %in% 2000:2009)

ggplot() +
  geom_sf(data = world_countries) +
  coord_sf(xlim = c(-110, 0), ylim = c(5, 65)) +
  geom_path(data = some_storms, color = "#1C3EFF", alpha = 0.9, 
          aes(x = long, y = lat, group = name)) +
  facet_wrap(~ year, nrow = 2) +
  theme(panel.background = element_blank(),
        axis.ticks = element_blank(), # hide tick marks
        axis.text = element_blank()) + # hide degree values of lat & lon
  labs(title = "North Atlantic Storms (2000-2009)",
       x = "", 
       y = "")
```


## Maps in R

Our first approach involves packages:

- `"tidyverse"` (data wrangling and exploration)

- `"sf"` (simple features)

- `"rnaturalearth"` (maps data)


## Main Idea

Assuming that you have an `"sf"` (simple features) object (i.e. map data),
you pass this to `ggplot()` and use the `geom_sf()` layer.

```{r}
#| eval: false
#| echo: true
# (abstract code)
ggplot() + 
  geom_sf(data = map_object)
```

. . .

<br>

Or also (with more customizing options)

```{r}
#| eval: false
#| echo: true
# (abstract code)
ggplot() + 
  geom_sf(data = map_object) +
  coord_sf(...)
```


## Map Data from `"rnaturalearth"`

Examples:

- `ne_coastline(returnclass = "sf")`

- `ne_countries(returnclass = "sf")`

- `ne_countries(country = "mexico", returnclass = "sf")`



## Coast Line World Map

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
world_coast = ne_coastline(returnclass = "sf")

ggplot() + 
  geom_sf(data = world_coast)
```


## Map of a given continent (e.g. North America)

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
map_north_america = ne_countries(
  continent = "north america", 
  returnclass = "sf")

ggplot() + 
  geom_sf(data = map_north_america) +
  theme(panel.background = element_blank())
```


## Map of a given country (e.g. Mexico)

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
map_mexico1 = ne_countries(country = "mexico", returnclass = "sf")

ggplot() + 
  geom_sf(data = map_mexico1) +
  theme(panel.background = element_blank())
```


## Map of a given country, with states (e.g. Mexico)

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
# (requires pkg "rnaturalearthhires")
# remotes::install_github("ropensci/rnaturalearthhires")
map_mexico2 = ne_states(country = "mexico", returnclass = "sf")

ggplot() + 
  geom_sf(data = map_mexico2) +
  theme(panel.background = element_blank())
```


## Adding Data to a Map

```{r}
#| eval: false
#| echo: true
# (abstract code)
ggplot() + 
  geom_sf(data = map_object) +
  geom_...(data = data_table, aes(...))
```

. . .

<br>

Or also:

```{r}
#| eval: false
#| echo: true
# (abstract code)
ggplot() + 
  geom_sf(data = map_object) +
  geom_...(data = data_table, aes(...)) +
  geom_...(data = data_table, aes(...)) 
```


## Adding Data to a Map

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
storms75 = storms |> filter(year == 1975)

ggplot() + 
  geom_sf(data = map_north_america) +
  geom_point(data = storms75, 
             aes(x = long, y = lat, group = name, color = name)) +
  theme(panel.background = element_blank()) +
  labs(title = "North Atlantic Storms (1975)")
```


## Adding Data to a Map

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
storms75 = storms |> filter(year == 1975)

ggplot() + 
  geom_sf(data = map_north_america) +
  geom_point(data = storms75, size = 0.5,
             aes(x = long, y = lat, group = name, color = name)) +
  geom_path(data = storms75, 
             aes(x = long, y = lat, group = name, color = name)) +
  theme(panel.background = element_blank()) +
  labs(title = "North Atlantic Storms (1975)")
```


## Adding Data to a Map

```{r}
#| code-fold: true
#| code-summary: View Code
#| echo: true
storms75 = storms |> filter(year == 1975)

ggplot() +
  geom_sf(data = world_countries) +
  coord_sf(xlim = c(-110, 0), ylim = c(5, 65)) +
  geom_path(data = storms75, 
          aes(x = long, y = lat, group = name, color = name)) +
  facet_wrap(~ year, nrow = 2) +
  theme(panel.background = element_blank(),
        axis.ticks = element_blank(), # hide tick marks
        axis.text = element_blank()) + # hide degree values of lat & lon
  labs(title = "North Atlantic Storms (1975)",
       x = "", 
       y = "")
```
